Structural Breaks in Inflation Dynamics within the European Monetary Union

### 2011-05-24 pkg/paper/Paper/draft.Rnw
rewrote intro to modeling section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/glogis@99 b679b2f7-6539-45d0-a605-5ddcb976cde9

To monitor changes in the dynamics of the inflation rates, we follow the usual
strategy employed when testing for structural breaks. First, the model is fitted
for the whole sample using the generalized logistic distribution. This then
enables testing for the stability of the parameter estimates. If there is
evidence of significant changes, the number and the timing of the break points
are estimated.

This framework allows to trace the evolution of mean, variance and skewness of
the inflation series and thus interpreting deviations in any of these parameters
as possibly due to a change in monetary policy, is made feasible. Any huge
intervention or change is captured by fluctuations in the parameter stability of
the series and can be quantified and related to a distinct time frame.

The correlation \fixme{Jesus: yes, but a DGP needs to be assumed, it would need
to take care of any autororrelation, wouldn't it?} dynamics of the series is of
less importance to this framework, which concentrates on changes in mean,
variance and skewness and does not focus on autocorrelation, the use of
alternative tests, as described in \cite{andreou}, is also hindered by the low
number of observations, due to the monthly frequencies.\footnote{In some years
the Billion Prices Project of the \cite{bpp}, which provides a much higher data
frequency, could solve this problem.}

>>>>

The goal of the modeling strategy, that is subsequently introduced, is the
following: detection of structural changes in the distributional properties of
the inflation rates over time, e.g., as potentially caused by interventions.
However, unlike many standard least squares approaches \citep[such as][]{bp03},
not only changes in the mean level should be addressed but also changes in
variance and potential skewness.

Hence, a likelihood-based model is adopted that can also incorporate stylized
facts for return data, such as heavy tails and skewness. Here, the seasonally
adjusted returns $y_t$ ($t = 1, \dots, n$) are assumed to come from a
generalized logistic (GL) distribution \citep[see][]{johnson} with
three-dimensional parameter $\phi_t = (\theta_t, \sigma_t, \delta_t)^\top$ (for
location~$\theta_t$, scale~$\sigma_t$, and shape~$\delta_t$ at time $t$). Structural change
techniques are then employed to check whether the parameter vector $\phi_t$
remains constant over time and, if that is not the case, when and how the
parameters change. This framework allows to trace breaks in the
evolution of mean, variance and skewness of the inflation series which may
possibly be linked to underlying regime changes, e.g., in monetary policy.

For estimation of the parameters, a (quasi-)maximum likelihood framework is
adopted and the observations are assumed to be (approximately) independent,
i.e., potential autocorrelation is treated as a nuisance parameter and not
explicitly incorporated in the model. The motivation for this is to focus on the
shifts in mean, variance, and skewness -- and it turns out to be a useful model
for the data under investigation because only negligible amounts of residual
autocorrelation remain after incorporation of the structural breaks. If the
focus of the analysis were the dynamics in the autocorrelation, higher frequency
data would be required, e.g., in order to apply techniques such as those
expounded by \cite{andreou}.\footnote{For example, it would be conceivable to
obtain high-frequency data for certain years from the Billion Prices Project of
the \cite{bpp}; however, this is not pursued here.}

\fixme{Z: potentially say something about the subsequent subsections}
